---
title: "Data Preprocessing and Raster Manipulation"
author: "Mark Baker"
output: pdf_document
---

# Importing Key Libraries
```{r}
library(raster)
library(sp)
library(sf)
library(rgdal)
library(leaflet)
library(ggplot2)
library(rasterVis)

```
# Loading in Raster Necessary Data
```{r}
cdl2019 <- raster("/Users/markbaker/Downloads/GEOG28602/final/2019_30m_cdls/2019_IL.tif")
cdl2019
```

```{r}
cdl2018 <- raster("/Users/markbaker/Downloads/GEOG28602/final/2018_30m_cdls/2018_IL.tif")
cdl2018
```

# Raster Manipulation

In this selection, I will work to generate the key output file that will utlized in the creation of the Shiny Application 

### Selecting Forestry Cover

```{r}
d2018 <- cdl2018 == 141 #Deciduous
e2018 <- cdl2018 == 142 #Evergreen
m2018 <- cdl2018 == 143 #Mixed
forest2018 <- d2018 + e2018 + m2018 #Total
plot(forest2018)
```

```{r}
d2019 <- cdl2019 == 141 #Deciduous
e2019 <- cdl2019 == 142 #Evergreen
m2019 <- cdl2019 == 143 #Mixed
forest2019 <- d2019 + e2019 + m2019 #Total
plot(forest2019)
```


### Raster Algebra
```{r}
c1819 <- -2*forest2018 + forest2019 + 3
plot(c1819)
```
### Displaying deforested, reforested, and areas of no change.
Confirming that the deforested, reforested, and areas of no change are apparent.

```{r}
df1819 <- c1819 == 1
nf1819 <- c1819 == 2
rf1819 <- c1819 == 4
plot(df1819)
plot(nf1819)
plot(rf1819 )
```

# Writing Raster Data to Computer for Shiny Application.
```{r}
writeRaster(c1819, filename="./change/c1819.tif", format="GTiff", overwrite=TRUE)
```

# Implementing Vector Layer
Here, I am reading in the Illinois County Shapefile for future manipulation.
```{r}
il <- st_read("/Users/markbaker/Downloads/GEOG28602/final/IL_BNDY_County/IL_BNDY_County_Py.shp")
il <- st_transform(il, "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
```

# Cropping Raster Data to Particular Regions (A Case Example)

Examining key avenues to examine when implementing crop feature for raster data. This will be heavily utilized when attempting to display particular counties in Illinois in the Shiny App.

### Cropping by Extent
By cropping by extent first, we will minimize the time it takes to process the raster when attempting to display the entire raster data.
```{r}
cook <- il[il$COUNTY_NAM == "COOK",]
p_cook_forest2018 <- intersect(c1819, cook)

```

### Creating a mask of the polygon of interest
Assigning value of 1 to every pixel within area of interest and 0 to other pixels.

```{r}
blank <- raster(p_cook_forest2018)
land_raster <- rasterize(cook, blank)
land_raster[!(is.na(land_raster))] <- 1
land_raster[(is.na(land_raster))] <- 0
plot(land_raster)
```

### Applying the Mask to the forestry cropped extent
Using this, we will reach the end goal of cropping the raster to a particular polygon used.
```{r}
cook_raster <- p_cook_forest2018 * land_raster
plot(cook_raster)
```

# Generating Land Cover Calculator
Below is a function that counts the total number of pixels in the forestry change raster file and outputs a well-formated data.frame.
```{r}
change_table <- function(raster) {
  freq_table <- freq(raster, digits=0, value=NULL)
  freq_table <- as.data.frame(freq_table)
  freq_table <- freq_table[(freq_table$value == 1| freq_table$value == 2 | 
                            freq_table$value == 3 | freq_table$value == 4) ,]
  rfreq <- head(freq_table[freq_table$value == 4, ],1)$count
  dfreq <- head(freq_table[freq_table$value == 1, ],1)$count
  nfreq <- head(freq_table[freq_table$value == 2, ],1)$count
  other <- head(freq_table[freq_table$value == 3, ],1)$count
  forest_calc <- data.frame("Type" = c("Reforested", "Deforested", "No Change in Forest", "Other Land"),
                          "Value" = c(4,1,2, 3),
                          "Percent" = c(rfreq/ sum(head(freq_table, 4)$count) *100,
                                        dfreq/ sum(head(freq_table, 4)$count) *100,
                                        nfreq/ sum(head(freq_table, 4)$count) *100,
                                        other/ sum(head(freq_table, 4)$count) *100))
  forest_calc
}

```

Here I am determine a simple way to remove the "Value" column from the generated table from change\_table function.
```{r}
table<- change_table(cook_raster)
table <- table[, -2]
table
```




