---
title: "Geographic Information Science III - Lab 1"
author: "Mark Baker"
output: pdf_document
---

### Verifying that verison of R:

```{r}
R.version.string
```

After running this code, it is clear that R version 3.6.0 was utilized.

## Examining Vector Data

### Installing and Loading Necessary Packages

Installing Packages
```{r}
# packages were installed and then commented 
#install.packages("sf")
#install.packages("raster")
#install.packages("spData")
```

Loading Libraries

```{r}
library(sf)          # classes and functions for vector data
library(raster)      # classes and functions for raster data
library(spData)        # load geographic data
```


### Examining the world spatial object

Plotting the World: 
```{r}
# the names of world can be returned using name()
names(world)

# the function plot() can be used for visualization
plot(world)
```

Being able to treat spatial objects as regular dataframes with spatial powers has advantages. For instance, we can summarize the life expectancies: 

```{r}
summary(world["lifeExp"])
```

Additionally, we are able to see \textbf{sf} objects are easy to subset as well.

```{r}
world_mini = world[1:2, 1:3]
world_mini
```

## TroubleShooting:
Working through this Lab, I did not run into any large problems.
