---
title: "Geographic Information Science III - Lab 2"
author: "Mark Baker"
output: pdf_document
---

## Loading Libraries:
```{r}
library(sf)
library(spData)
```

## The Dataset of Interest:
For this lab, I will be using the $world$ dataset that is in the $spData$ package
```{r}
class(world)
```
Looking at the class of the world dataset, we can see that it is loaded as a "data.frame" in simple features as desired.

## Summary of Key Variables in World Dataset:
```{r}
summary(st_drop_geometry(world[ , c("area_km2", "pop", "lifeExp", "gdpPercap")]))
```

Looking at the summary table, we see that the mean area is $832558 km^2$ for all countries, the mean population is $4.282e+07$ people, the mean life expectancy is $70.85$ years, and the mean GDP per capita is $17106.0$.
\newline
Based on the findings above, future analysis will focus on GDP per capita.
## GDP per Capita
```{r, fig.height = 4}
hist(world$gdpPercap, main = "Distribution of GDP per capita for all Countries",
     ylim=c(0, 80), xlab = "GDP per capita", col = "gray")
```

Based on the histogram above, it is clear that a majority of countries have GDP per capita less than $\$20,000$.

```{r, fig.height = 4}
plot(world[, "gdpPercap"], main = "GDP per Capita for all countries")
```
\newline
Specifically examining the generated map, we see that areas with higher GDP per capita are located in North America, Europe, the Middle East, and Austrailia.
